A Decision Procedure for (Co)datatypes in SMT Solvers

### 2015-02-19 tex/body.tex
more

First, we need the following definitions.
We write $t \vsimv{x} u$ if $\mu$-terms $t$ and $u$ are syntactically equivalent
for some renaming that avoids capturing any variable other than $x$.
For example,
$\MU x.\; \const{D}( x ) \vsimv{y} \MU x.\; \const{D}( y )$
(by renaming $y$ to $x$),
$\MU x.\; \const{C}( x, x ) \vsimv{x} \MU y.\; \const{C}( x, y )$, and
$\MU x.\; \const{C}( z, x ) \vsimv{z} \MU y.\; \const{C}( z, y )$,
but
$\MU x.\; \const{D}( x ) \not\vsimv{x} \MU x.\; \const{D}( y )$ and
$\MU x.\; \const{C}( x, x ) \not\vsimv{y} \MU y.\; \const{C}( x, y )$.
For a variable $x\typ{\tau}$ and a normal interpretation $\J$,
we write $\Terms^{\smash{\J}}_x( \Val )$ to denote the set consisting of all values
$v \in \J( \tau )$
such that $v \vsimv{x} \interp{t}{u}$ for some term $t$ with subterm $u$ occurring in the range of $\Val$.

@@

\begin{examplex}
Let $\delta$ be a codatatype with the constructors $\const{C},\, \const{D},\, \const{E} :\delta \rightarrow \delta$.
Let $\Ec$ be the set
$\{
u \teq \const{C}( z ),\;
v \teq \const{D}( z ),\;
w \teq \const{E}( y ),\;
x \teq \const{C}( v ),\;
z \not\teq v
\}$.
After applying the calculus to saturation on $\Ec$, the mapping $\Val$ is as follows:
%\[\begin{array}{r@{}c@{}l@{\;}l@{\qquad}r@{}c@{}r@{\;}l@{}l@{\;}l}
%\[\begin{array}{r@{}c@{}l@{\;}l@{\qquad}r@{}c@{}r@{\;}l@{}l@{\;}l@{\qquad}r@{}c@{}r@{\;}l}
\[\begin{array}{r@{}c@{}r@{}l@{\qquad}r@{}c@{}r@{}l@{\qquad}r@{}c@{}l}
\Val \ec{u} & {}\,=\,{} & \MU {\Varec{u}}.\; & \const{C}( {\Varec{z}} ) &
\Val \ec{w} & = & \MU {\Varec{w}}.\; & \const{E}( {\Varec{y}} ) &
\Val \ec{y} & {}\,=\,{} & {\Varec{y}} \\
\Val \ec{v} & = & \MU {\Varec{v}}.\; & \const{D}( {\Varec{z}} ) &
\Val \ec{x} & {}\,=\,{} & \MU {\Varec{x}}.\; & \const{C}( \MU {\Varec{v}}.\; \const{D}( {\Varec{z}} ) ) &
\Val \ec{z} & = & {\Varec{z}}
\end{array}\]
%
To construct a completion $\ValC$, we must choose values for the free
variables of $\Val$, specifically ${\Varec{y}}$ and ${\Varec{z}}$.
Modulo $\alpha$-equivalence, $\smash{\Terms^{\smash{\J}}_{{\Varec{z}}}( \Val )} =
%\begin{array}{c}
\{
\MU a.\: \const{D}( a ),\:
\MU a.\: \const{C}( \const{D}( a )),\allowbreak\:
\const{C}( \MU a.\: \const{D}( a )),\allowbreak\:
\MU a.\: \const{C}( a )
\}
%\end{array}
$.
Consider a normal interpretation $\J$ that evaluates variables in $\Ec$ based on the mapping $\Val$:
$z$ is interpreted as $\Val \ec{z}$, $y$ as $\Val \ec{y}$, and so on.
Assigning a value for $\Val \ec{z}$ that is $\alpha$-equivalent to a value in $\Terms^{\smash{\J}}_{{\Varec{z}}}( \Val )$
may cause values in the range of $\Val$ to become $\alpha$-equivalent,
which in turn may cause $\Ec$ to be falsified by $\J$.
For example, assign $\MU {\Varec{z}}.\; \const{D}( {\Varec{z}} )$ for ${\Varec{z}}$.
After the substitution, $\Val \ec{v} = \MU {\Varec{v}}.\; \const{D}( \MU {\Varec{z}}.\; \const{D}( {\Varec{z}} ) )$,
which has normal form $\MU {\Varec{v}}.\; \const{D}( {\Varec{v}} )$,
which is $\alpha$-equivalent to $\MU {\Varec{z}}.\; \const{D}( {\Varec{z}} )$.
However, this contradicts the disequality $z \not\teq v$ in~$\Ec$.
On the other hand, if the value assigned to ${\Varec{z}}$ is fresh, %not $\alpha$-equivalent to any term in $\Terms^{\smash{\M}}_{{\Varec{z}}}( \Val )$,
the values in the range of $\Val$ remain $\alpha$-disequivalent.
For example, we may assign a value such as
$\MU {\Varec{z}}.\; \const{E}( {\Varec{z}} )$ or
$\MU {\Varec{z}}.\; \const{D}( \const{C}( {\Varec{z}} ) )$
to ${\Varec{z}}$.
Legal substitutions for ${\Varec{z}}$ may cause the range of $\Val$ to contain abnormal terms.
For example, after assigning $\MU {\Varec{z}}.\; \const{D}( \const{C}( {\Varec{z}} ) )$ to $\Varec{z}$, we have
$\Val \ec{u} = \MU {\Varec{u}}.\; \const{C}( \MU {\Varec{z}}.\; \const{D}( \const{C}( {\Varec{z}} ) ) )$,
with normal form
$\MU {\Varec{u}}.\; \const{C}( \MU {\Varec{z}}.\; \const{D}( {\Varec{u}} ) )$.
\xend
\end{examplex}

>>>>

First, we need the following definitions.
We write $t \vsimv{x} u$ if $\mu$-terms $t$ and $u$ are syntactically equivalent
for some renaming that avoids capturing any variable other than $x$.
For example,
$\MU x.\; \const{D}( x ) \vsimv{y} \MU x.\; \const{D}( y )$
(by renaming $y$ to $x$),
$\MU x.\; \const{C}( x, x ) \vsimv{x} \MU y.\; \const{C}( x, y )$, and
$\MU x.\; \const{C}( z, x ) \vsimv{z} \MU y.\; \const{C}( z, y )$,
but
$\MU x.\; \const{D}( x ) \not\vsimv{x} \MU x.\; \const{D}( y )$ and
$\MU x.\; \const{C}( x, x ) \not\vsimv{y} \MU y.\; \const{C}( x, y )$.
For a variable $x\typ{\tau}$ and a normal interpretation $\J$,
we let $\Terms^{\smash{\J}}_x( \Val )$ denote the set consisting of all values
$v \in \J( \tau )$
such that $v \vsimv{x} \interp{t}{u}$ for some subterm $u$ of a term $t$
occurring in the range of $\Val$. This set describes shapes of terms to avoid.

@@

\begin{examplex}
Let $\delta$ be a codatatype with the constructors $\const{C},\, \const{D},\, \const{E} :\delta \rightarrow \delta$.
Let $\Ec$ be the set
$\{
u \teq \const{C}( z ),\;
v \teq \const{D}( z ),\;
w \teq \const{E}( y ),\;
x \teq \const{C}( v ),\;
z \not\teq v
\}$.
After applying the calculus to saturation on $\Ec$, the mapping $\Val$ is as follows:
%\[\begin{array}{r@{}c@{}l@{\;}l@{\qquad}r@{}c@{}r@{\;}l@{}l@{\;}l}
%\[\begin{array}{r@{}c@{}l@{\;}l@{\qquad}r@{}c@{}r@{\;}l@{}l@{\;}l@{\qquad}r@{}c@{}r@{\;}l}
\[\begin{array}{r@{}c@{}r@{}l@{\qquad}r@{}c@{}r@{}l@{\qquad}r@{}c@{}l}
\Val \ec{u} & {}\,=\,{} & \MU {\Varec{u}}.\; & \const{C}( {\Varec{z}} ) &
\Val \ec{w} & = & \MU {\Varec{w}}.\; & \const{E}( {\Varec{y}} ) &
\Val \ec{y} & {}\,=\,{} & {\Varec{y}} \\
\Val \ec{v} & = & \MU {\Varec{v}}.\; & \const{D}( {\Varec{z}} ) &
\Val \ec{x} & {}\,=\,{} & \MU {\Varec{x}}.\; & \const{C}( \MU {\Varec{v}}.\; \const{D}( {\Varec{z}} ) ) &
\Val \ec{z} & = & {\Varec{z}}
\end{array}\]
%
To construct a completion $\ValC$, we must choose values for the free
variables of $\Val$, specifically ${\Varec{y}}$ and ${\Varec{z}}$.
Modulo $\alpha$-equivalence, $\smash{\Terms^{\smash{\J}}_{{\Varec{z}}}( \Val )} =
%\begin{array}{c}
\{
\MU a.\: \const{C}( a ),\:
\MU a.\: \const{D}( a ),\allowbreak\:
\MU a.\: \const{C}( \const{D}( a )),\allowbreak\:
\const{C}( \MU a.\: \const{D}( a ))
\}
%\end{array}
$.
Consider a normal interpretation $\J$ that evaluates variables in $\Ec$ based on the mapping $\Val$:
$z$ is interpreted as $\Val \ec{z}$, $y$ as $\Val \ec{y}$, and so on.
Assigning a value for $\Val \ec{z}$ that is $\alpha$-equivalent to a value in $\Terms^{\smash{\J}}_{{\Varec{z}}}( \Val )$
may cause values in the range of $\Val$ to become $\alpha$-equivalent,
which in turn may cause $\Ec$ to be falsified by $\J$.
For example, assign $\MU {\Varec{z}}.\; \const{D}( {\Varec{z}} )$ for ${\Varec{z}}$.
After the substitution, $\Val \ec{v} = \MU {\Varec{v}}.\; \const{D}( \MU {\Varec{z}}.\; \const{D}( {\Varec{z}} ) )$,
which has normal form $\MU {\Varec{v}}.\; \const{D}( {\Varec{v}} )$,
which is $\alpha$-equivalent to $\MU {\Varec{z}}.\; \const{D}( {\Varec{z}} )$.
However, this contradicts the disequality $z \not\teq v$ in~$\Ec$.
On the other hand, if the value assigned to ${\Varec{z}}$ is fresh, %not $\alpha$-equivalent to any term in $\Terms^{\smash{\M}}_{{\Varec{z}}}( \Val )$,
the values in the range of $\Val$ remain $\alpha$-disequivalent.
For example, we may assign a value such as
$\MU {\Varec{z}}.\; \const{E}( {\Varec{z}} )$ or
$\MU {\Varec{z}}.\; \const{D}( \const{C}( {\Varec{z}} ) )$
to ${\Varec{z}}$.
Legal substitutions for ${\Varec{z}}$ may cause the range of $\Val$ to contain abnormal terms.
For example, after assigning $\MU {\Varec{z}}.\; \const{D}( \const{C}( {\Varec{z}} ) )$ to $\Varec{z}$, we have
$\Val \ec{u} = \MU {\Varec{u}}.\; \const{C}( \MU {\Varec{z}}.\; \const{D}( \const{C}( {\Varec{z}} ) ) )$,
with normal form
$\MU {\Varec{u}}.\; \const{C}( \MU {\Varec{z}}.\; \const{D}( {\Varec{u}} ) )$.
\xend
\end{examplex}
